---
title: "Lotka Volterra Competiton"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
lv <- function(time,init,params) {
  with (as.list(c(time,init,params)), {
    # description of parameters:
    # r1 = growthrate of Sp. 1; r2 = growthrate of Sp. 2
    # N = population size of Sp. 1; P = Population Size Sp. 2
    # a = competitive impact of Sp. 2 on Sp. 1; b = competitive impact of Sp 1 on Sp 2
    # K1/K2 = carrying capacities
    dN1dt <- (r1*N1*(1-(N1+a*N2)/K1)) 
    dN2dt <- (r2*N2*(1-(N2+b*N1)/K2))
    return(list(c(dN1dt,dN2dt)))
  })
}

# Root functions define the stopping parameters - when this root is zero, the thing will stop.
rootfun <- function(Time, State, Pars) {
  dstate <- unlist(lv(Time, State, Pars))
  sum(abs(dstate)) - 1e-3
}

```

## The Lotka-Volterra Competition Model

The most basic Lotka-Volterra competition models describe population dynamics of a pair of competing species. In the absense of competition, both species are predicted to grow according to a [logistic growth model](https://en.wikipedia.org/wiki/Logistic_function) to some carrying capacity $K$. 

In the competition equations, the interaction parameters $\alpha$ and $\beta$ describe the influence that one individual of species $i$ has on the population growth rate of species $j$. 

I suggest that you get a solid intuition for the assumptions behind Lotka Volterra competition before using this app. Nick Gotelli's [Primer on Ecology](http://www.sinauer.com/a-primer-of-ecology.html) is a fantastic resource, but any basic ecology textbook should do. There's also a lot of useful resources available online- just google away!

### Model description

These are the equations that describe Lotka-Volterra competition dynamics:   
$$\frac{dN_{1}}{dt} = (r_{1}N_{1}) \left( 1-\frac{N_{1} + \alpha*N_{2}}{K_{1}} \right)$$
$$\frac{dN_{2}}{dt} = (r_{2}N_{2}) \left( 1-\frac{N_{2} + \beta*N_{1}}{K_{2}} \right)$$

We are generally interested in finding out when two competing species can *coexist* in the long term. To do this, we need to find out what the *long term equilibrium dynamics* of the model are. Because populations don't grow at equilibrium (by definition), we can set the population growth equations equal to zero and solve for values of $N$s. You should do the math independently, and verify that you get the following solution:

$$\frac{dN_{1}}{dt} = 0 \text{  when  } N_{1} = K_{1} - \alpha*N_{2}$$
$$\frac{dN_{2}}{dt} = 0 \text{  when  } N_{2} = K_{2} - \beta*N_{1} $$


## Simulating the dynamics of the model

```{r eruptions, echo=FALSE, fig.height=30}

sidebarPanel(
  h4("Set initial population sizes"), br(),
  fluidRow(
  column(6, numericInput("n1",label=p("Set initial", span("N1",style = "color:blue")), value = floor(rnorm(1,mean=500,sd=100)),min=1)),
  column(6, numericInput("n2",label=p("Set initial", span("N2",style = "color:red" )), value = ceiling(rnorm(1,mean=500,sd=100)),min=1))),

  h4("Set carrying capacities"),
  fluidRow(
    column(6, numericInput("k1",label=p("Set", span("K1",style = "color:blue")), value = 500, min = 1)),
    column(6, numericInput("k2",label=p("Set", span("K2",style = "color:red" )), value = 500, min = 1))),
  
  h4("Set competition coefficients"),
  fluidRow(
  column(6, sliderInput("alpha", 
              label = p("Set", HTML("&alpha;")),
              min = 0, max = 2.0, value=0.75, step = 0.01)),
  column(6, sliderInput("beta", 
              label = p("Set", HTML("&beta;")),
              min = 0, max = 2.0, value=0.75, step = 0.01))),
  
  br(),
  radioButtons("iter", label="",
               choices = list("Run until steady state" = 1,
                              "Run for specified length" = 2),
               selected = 1),
  htmlOutput("UI1"),
  downloadButton("downloadPlot", "Download Plot!")
  
)

output$UI1 <- renderUI({
  if (input$iter == 2){
    numericInput("steps",label=h6("Time steps to run the model"),value=100,min=1)}
  else {""}
})

init <- reactive ({
  c("N1" = input$n1, "N2" = input$n2)
})

# Reactively generate the params list. Access later using params()
params <- reactive({
  c("r1" = 1, "r2" = 1, "K1" = as.numeric(input$k1), "K2" = as.numeric(input$k2), "a" = as.numeric(input$alpha), "b" = as.numeric(input$beta))
})

# Reactively generate the time. Access later using time()
time <- reactive({
  
  if (input$iter == 2){seq(0,input$steps,by=1)}
  else {seq(0,1000, by=1)}
})

# Reactively generate the lvout dataframe. Access later using lvout()
lvout <- reactive({
  init<- init()
  params <- params()
  time <- time()
  
  if (input$iter == 2){floor(as.data.frame(ode(func=lv,y=init,parms=params,times=time)))}
  else {floor(as.data.frame(lsodar(func=lv,y=init,parms=params,times=time,rootfun=rootfun)))}
})
# Generate the plots
mainPanel(renderPlot({
  init<- init()
  params <- params()
  time <- time()
  generated_df <- lvout()      
  
  par(mfrow=c(2,2))
  # plot the Zero Net Growth Isoclines based on parameters above.
  plot(1,type="n",xlim=c(0,max(params["K1"],params["K2"]/params["b"])*1.25),
       ylim=c(0,max(params["K2"],params["K1"]/params["a"])*1.25),
       xlab  ="Species 1",ylab="Species 2",main="ZNGIs for Sp.1 and Sp.2",
       xaxs="i",yaxs="i",cex.axis=1.5,cex.lab=1.5,cex.main=1.5)
  legend("topright", col = c("Blue", "Red"), lty = 1, legend = c("Species 1", "Species 2"), bty = "n")
  lines(x = c(params["K1"],0),y = c(0,params["K1"]/params["a"]),lwd=2,col="blue")
  lines(x = c(params["K2"]/params["b"], 0) ,y = c(0,params["K2"]),lwd=2,col="red")
  
  # plot the starting population size
  points(x=init["N1"],y=init["N2"],cex=2,pch=20)
  
  # Plot out the results of the ODE. 
  # First plot is for N vs P (sp1 v sp2)
  plot(generated_df$N2~generated_df$N1,type="o",xlim=c(0,max(params["K1"],params["K2"]/params["b"])*1.25),
       ylim=c(0,max(params["K2"],params["K1"]/params["a"])*1.25), main = "Trajectory of population sizes",xlab="Species 1", 
       ylab="Species 2",cex.axis=1.5,cex.lab=1.5,cex.main=1.5, col=rainbow(nrow(generated_df)))
  
  # Second plot for N & P v Time
  plot(generated_df$N1,ylim=c(0,max(max(generated_df$N1),max(generated_df))*1.25),ylab="Population size",
       type="l",lwd=2, col="blue",cex.axis=1.5,cex.lab=1.5,
       main="Population size vs time",cex.main=1.5)
  points(generated_df$N2,col="red",type="l",lwd=2)
  legend("topright", col = c("Blue", "Red"), lty = 1, legend = c("Species 1", "Species 2"), bty = "n")      
}, height = 800, width = 700))

plot2save <- function(){
  init<- init()
  params <- params()
  time <- time()
  generated_df <- lvout()      
  
  par(mfrow=c(2,2))
  # plot the Zero Net Growth Isoclines based on parameters above.
  plot(1,type="n",xlim=c(0,max(params["K1"],params["K2"]/params["b"])*1.25),
       ylim=c(0,max(params["K2"],params["K1"]/params["a"])*1.25),
       xlab  ="Species 1",ylab="Species 2",main="ZNGIs for Sp.1 and Sp.2",
       xaxs="i",yaxs="i",cex.axis=1.5,cex.lab=1.5,cex.main=1.5)
  legend("topright", col = c("Blue", "Red"), lty = 1, legend = c("Species 1", "Species 2"), bty = "n")
  lines(x = c(params["K1"],0),y = c(0,params["K1"]/params["a"]),lwd=2,col="blue")
  lines(x = c(params["K2"]/params["b"], 0) ,y = c(0,params["K2"]),lwd=2,col="red")
  
  # plot the starting population size
  points(x=init["N1"],y=init["N2"],cex=2,pch=20)
  
  # Plot out the results of the ODE. 
  # First plot is for N vs P (sp1 v sp2)
  plot(generated_df$N2~generated_df$N1,type="o",xlim=c(0,max(params["K1"],params["K2"]/params["b"])*1.25),
       ylim=c(0,max(params["K2"],params["K1"]/params["a"])*1.25), main = "Trajectory of population sizes",xlab="Species 1", 
       ylab="Species 2",cex.axis=1.5,cex.lab=1.5,cex.main=1.5, col=rainbow(nrow(generated_df)))
  
  # Second plot for N & P v Time
  plot(generated_df$N1,ylim=c(0,max(max(generated_df$N1),max(generated_df))*1.25),ylab="Population size",
       type="l",lwd=2, col="blue",cex.axis=1.5,cex.lab=1.5,
       main="Population size vs time",cex.main=1.5)
  points(generated_df$N2,col="red",type="l",lwd=2)
  legend("topright", col = c("Blue", "Red"), lty = 1, legend = c("Species 1", "Species 2"), bty = "n")      
}

param_text <- reactive({
  init<- init()
  params <- params()
  
  paste("K1 = ", params["K1"], ", K2 = ", params["K2"], ", r1 = ", params["r1"], ", r2 = ", params["r2"], ", Alpha = ", params["a"], ", Beta = ", params["b"], sep = "")
})

output$downloadPlot <- downloadHandler(
  filename = 
    function() {
      paste('plot-', Sys.Date(), '.png', sep='')
    },
  content = function(filename) {
    png(filename, height = 500, width = 750)
    par(mfrow = c(1,2), oma = c(2,0,0,0))
    plot2save()
    mtext(param_text(), side = 1, outer = TRUE)
    dev.off()
  }
)
```
    
    
    
*This app was last updated on `r date()`*    
*Created and maintained by [Gaurav Kandlikar](http://gaurav.kandlikar.com) (gkandlikar@ucla.edu)*